\Huge{Vipassana for Hackers} \\ \large{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-10-04 paper-two/paper-two.tex
reviews of second edit with preethical

\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high
  level.}

@@

The regions involved in an individual's conceptual ``self'' are the Posterior
Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
volunteers were asked to reflect on personal preferences, beliefs, abilities, or
any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
blood flow and metabolic rate are 40\% higher than average in the PCC,
making it one of the most active regions of the brain. \cite{pccrole}

The PCC has been thought to mediate interactions between emotion and memory.
Under an fMRI, it consistently lights up when a person recollects an event from their
life. The strength of this activity varies depending on the emotional importance of
the memory recalled. It is also activated by emotional stimuli,
both positive and negative, and acts as an navigational interface, between the individual
and the external world, by gauging the importance of different stimuli. \cite{pccemotion}

The vmPFC is involved in creating a conceptual self by self-related processing
and the assignment of personal significance to self-related
information. \cite{dmnself} The vmPFC is paired with the Dorsomedial Prefrontal
Cortex (dmPFC), which assists in the creation of the conceptual self through an
individual's consideration of ``others''.

@@

\todo{Expand on IPL: Why is facial recognition interesting to this paper?}

It is hypothesized that the activation of these three regions of the brain (PCC,
vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
in space and time. It functions as a network in which phenomena such as self
representations are accessible via the PCC \cite{mappingself} and are gated into
conscious awareness by activity in the mPFC \todo{The mPFC needs to be mentioned
  earlier, defined, and abbreviated.} , as influenced by changing internal
and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
created by a large scale integration of DMN activity. \cite{mappingself}

@@

\todo{Define and describe the Medial Temporal Subsystem: Describe MTL anatomically and MTS relation
  to the DMN, particularly the partial inclusion of the LS. Use the Anatomical
  introduction (immediately following the Introduction) to assist in framing this.}

@@

The critical role of the MTL in long term memory is well documented. \cite{mtl}
This network \todo{Which network?} has also been activated during autobiographical memory retrieval,
contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}

@@

When a meditator sits down to meditate and thoughts wander away to past experiences
or anxieties about the future, it is the DMN playing the most active role in this
distraction.

\todo{Expand on DMN relationship to Monkey Mind. Add papers with supporting evidence?}

@@

\subsubsection{Posterior Parietal Cortex}
The PPC is a region of the Parietal Cortex, physically located behind the primary
Somatosensory Cortex. The PPC receives input from auditory, visual, and somatosensory systems
as well as a variety of other brain regions \todo{What other brain regions provide
  input to the PPC?}, and it integrates this input to facilitate
the execution of functions that require diverse information (also known as higher-order functions).
In turn, the PPC activates the DL-PFC and motor cortex and is involved in the performance of
attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}

\todo{PSC section mentions Parietal Lobe but it hasn't been mentioned earlier, nor defined?}

\subsubsection{Sidenote: Somatosensory Cortex}
The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Lobe.
It is responsible for processing somatic sensations. These sensations arise from receptors
positioned throughout the body. They are responsible for detecting touch, for proprioception
(the position of the body in space), nociception (pain), and temperature. \cite{somato}
When such receptors detect one of these sensations, the information is sent first to
the Thalamus and then to the Primary Somatosensory Cortex.

@@

The Central Executive Network is involved in goal-directed behavior and processes
related to attention. These goals are usually external and attention is also directed
externally (regardless of whether attention is voluntary or not). When you meditate,
attention is directed both internally and voluntarily, to the same sensory cues that
are normally directing the PPC. \todo{What is interesting about internal/voluntary
  attention in CEN?}

@@

The SN is directly involved in switching between the DMN and CEN. When you meditate,
you cultivate the cognitive processes related to attention.
When your mind wanders, you notice it and bring it back.
The SN is directly involved in this process of assigning attention, since it is
responsible for the prominence of any object of attention.
Meditation, or the act of returning to an object of meditation, can therefore be
thought of as an exercise in strengthening the SN and its ability to switch from
the default objects of attention, over time.

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days students are asked
to focus their attention on the area below their nostrils and above their upper
lip. That's all. These are undoubtedly very simple instructions but in practice I
found them very difficult. Maintaining my focus was much harder than I had
anticipated. Within the first two days, I had made up elaborate stories about my
fellow meditation students. In these fantasies all of the other students were
superheroes, tirelessly working to save humanity together. My Default Mode Network
was in overdrive.

@@

\todo{Elaborate on breath/bridge and how it relates to brain function?}

@@

As I gave in to this activity, tried to focus more intently, and tried to sit still for
longer continuous stretches of time, there were longer periods of awareness on that
patch of skin. With these longer periods of awareness came
stranger sensations arising and passing within that physical area. By the third day,
it felt like entire ecosystems were writhing and flopping and crashing, all of them
very alive in the area below my nostrils above my upper lip.

\todo{Neurological / physiological basis for these strange sensations?}

@@

\todo{Relate ``gateway to altered states of mind'' to the Anapana ``bridge''?}

There has also been research in mice, showing a cluster of nerves called the
pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
with every breath taken. This breathing pacemaker seems to work not only for regular
conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
sighs and gasps. The preBötC also appears to play a role in calming and
arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
these neuron clusters during the first three days of a Vipassana course, even among
novice meditators.

@@

But Vipassana is an internal task. The Salience Network is activated by focused
attention, which in turn activates the Central Executive Network. Rather than an
external task, as the CEN is accustomed to, the conscious instruction is to observe
bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
not moving your body physically but instead moving throughout your body with your
attention, the CEN is doing what it has always done. However, the outcomes of
activity in the CEN is different during Vipassana. Instead of focusing on
external sensory input the sensory input is internal (bodily sensations) and instead
of reacting to input by performing tasks or feeling emotions the goal is to not
react, to simply observe as objectively as possible.

@@

Over the final seven days of the course, as awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. While
continuing to pay attention to constantly changing bodily sensations the goal remains: avoid
assigning evaluations (and therefore, emotions) to them. Without assigning an
emotional response, one learns to observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, reactions become smaller and less
frequent. This snowball effect almost mirrors the spiral of brain behaviours
exhibited early in the course. Early in the course, the CEN's normal mode of
operation is in play, which is to experience unpleasant stimuli and react (desiring to rid the body of
them), which in turn causes the unpleasant stimuli to appear all the more
unpleasant. Once a measure of objectivity is reached, however, the brain's function
snowballs in the other direction. It is as if both conscious and unconscious awarenesses are slowly aligning
until there are occasions of experience when it feels as if there is nothing but detached awareness moving
through the body (or some parts of the body). This is not at all akin to the
detachment of sleep or intoxication, however, and the senses are not dulled. In
addition to a feeling of emotional detachment, the Insula and the Somatosensory
Cortex, which are involved in sensory awareness, were \textbf{more} active, not
less. As the ACC, OFC, and Limbic System are calmed and attention is narrowed
further, the sensations become progressively more intense.

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the Insula
fires more rapidly, leading to alternative networks of consciousness
arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the ten
day course is well-designed \todo{How is the course well designed?} and will give someone experimenting with meditation the
minimum observational change necessary to instill an interest, to keep practicing.

>>>>

@@

The regions involved in an individual's conceptual ``self'' are the Posterior
Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
subjects were asked to reflect on personal preferences, beliefs, abilities, or
any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
blood flow and metabolic rate are 40\% higher than average in the PCC,
making it one of the most active regions of the brain. \cite{pccrole}

The PCC has been thought to mediate interactions between emotion and memory.
Under an fMRI, it consistently lights up when a person recollects an event from their
life. The strength of this activity varies depending on the emotional importance of
the memory recalled. It is also activated by emotional stimuli,
both positive and negative, and acts as a navigational interface, between the individual
and the external world, by gauging the importance of different stimuli. \cite{pccemotion}

The vmPFC is involved in creating a conceptual self by self-related processing
and the assignment of personal significance to self-related
information. \cite{dmnself} The vmPFC is paired with the Dorsomedial Prefrontal
Cortex (dmPFC), which assists in the creation of the conceptual self through an
individual's consideration of ``others''. Together, the vmPFC and dmPFC form the
Medial Prefrontal Cortex (mPFC).

@@

\todo{Expand on IPL: How does the IPL contribute to our sense of self?}

It is hypothesized that the activation of these three regions of the brain (PCC,
vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
in space and time. It functions as a network in which phenomena such as self
representations are accessible via the PCC \cite{mappingself} and are gated into
conscious awareness by activity in the mPFC, as influenced by changing internal
and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
created by a large scale integration of DMN activity. \cite{mappingself}

@@

\todo{Define and describe the MTS: Describe MTL anatomically and MTS relation
  to the DMN, particularly the partial inclusion of the LS.}

@@

The critical role of the MTL in long term memory is well documented. \cite{mtl}
The MTS has also been activated during autobiographical memory retrieval,
contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}

@@

When a meditator sits down to meditate and thoughts wander away to past experiences
or anxieties about the future, classic meditation texts and instructions refer to
this behaviour as ``Monkey Mind''. It is the DMN playing the most active role in this
distraction.

@@

\subsubsection{Posterior Parietal Cortex}
The PPC is a region of the Parietal Cortex, physically located behind the primary
Somatosensory Cortex. The PPC is responsible for facilitating higher-order
functions. Higher-order functions are defined as those requiring diverse information
for the purposes of higher evolutionary behaviour: intelligence, memory, planning,
speech, orientation, decision-making, etc.

To do this, the PPC receives input from auditory, visual, and somatosensory
systems, integrates this input, and uses the aggregate to activate the DL-PFC and motor cortex and is
involved in the performance of attention related tasks as well as higher-order motor
tasks such as grasping and catching. \cite{parietallobesa,parietallobes}

\subsubsection{Sidenote: Somatosensory Cortex}
The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Cortex.
It is responsible for processing somatic sensations. These sensations arise from receptors
positioned throughout the body. They are responsible for detecting touch, for proprioception
(the position of the body in space), nociception (pain), and temperature. \cite{somato}
When such receptors detect one of these sensations, the information is sent first to
the Thalamus and then to the Primary Somatosensory Cortex.

@@

The CEN is involved in goal-directed behavior and processes
related to attention. These goals are usually external and attention is also directed
externally (regardless of whether attention is voluntary or not). When you meditate,
attention is directed both internally and voluntarily, to the same sensory cues that
are normally directing the PPC.

@@

The SN is directly involved in switching between the DMN and CEN. When you meditate,
you cultivate the cognitive processes related to attention.
When the meditator's mind wanders, she notices it and brings it back.
The SN is directly involved in this process of assigning attention, since it is
responsible for the prominence of any object of attention.
Meditation, or the act of returning to an object of meditation, can therefore be
thought of as an exercise in strengthening the SN and its ability to perform this switch, over time.

@@

And so it was. With my brain filled with thoughts about myself, about the people
around me, I started this meditation business. For the first 3.5 days I was instructed
to focus my attention on the area below my nostrils and above my upper
lip. That's all. These are undoubtedly very simple instructions but in practice I
found them very difficult. Maintaining my focus was much harder than I had
anticipated. Within the first two days, I had made up elaborate stories about my
fellow meditation students. In these fantasies all of the other students were
superheroes, tirelessly working to save humanity together. My Default Mode Network
was in overdrive.

@@

There has been research in mice, showing a cluster of nerves called the
pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
with every breath taken. This breathing pacemaker seems to work not only for regular
conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
sighs and gasps. The preBötC also appears to play a role in calming and
arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
these neuron clusters during the first three days of a Vipassana course, even among
novice meditators.

@@

As I gave in to this activity, tried to focus more intently, and tried to sit still for
longer continuous stretches of time, there were longer periods of awareness on that
patch of skin. With these longer periods of awareness came
stranger sensations arising and passing within that physical area. By the third day,
it felt like entire ecosystems were writhing and flopping and crashing, all of them
very alive in the area below my nostrils above my upper lip. Since Anapana meditation
is very simple and these strange sensations appeared within a few days of my practice
as an absolute beginner, it would be very interesting (and relatively easy) to study
the neurological and physiological basis for such sensations.

@@

@@

But Vipassana is an internal task. The Salience Network is activated by focused
attention, which in turn activates the Central Executive Network. Rather than an
external task, as the CEN is accustomed to, the conscious instruction is to observe
bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
not moving your body physically but instead moving throughout your body with your
attention, the CEN is doing what it has always done. However, the outcomes of
activity in the CEN are different during Vipassana. Instead of focusing on
external sensory input the sensory input is internal (bodily sensations) and instead
of reacting to input by performing tasks or feeling emotions the goal is to not
react, to simply observe as objectively as possible.

@@

Over the final seven days of the course, as awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. While
continuing to pay attention to constantly changing bodily sensations the goal remains: avoid
assigning evaluations (and therefore, emotions) to them. Without assigning an
emotional response, one learns to observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, reactions become smaller and less
frequent. This snowball effect almost mirrors the spiral of brain behaviours
exhibited early in the course. Early in the course, the CEN's normal mode of
operation is in play, which is to experience unpleasant stimuli and react (desiring to rid the body of
them), which in turn causes the unpleasant stimuli to appear all the more
unpleasant. Once a measure of objectivity is reached, however, the brain's function
snowballs in the other direction. It is as if both conscious and unconscious awarenesses are slowly aligning
until there are occasions of experience when it feels as if there is nothing but detached awareness moving
through the body (or some parts of the body). This is not at all akin to the
detachment of sleep or intoxication, however, and the senses are not dulled. In
addition to a feeling of emotional detachment, the AIC and the Somatosensory
Cortex, which are involved in sensory awareness, were \textbf{more} active, not
less. As the ACC, OFC, and Limbic System are calmed and attention is narrowed
further, the sensations become progressively more intense.

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the AIC
fires more rapidly, leading to alternative networks of consciousness
arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the
10-day course is both progressive and systematic. It is designed to give someone
experimenting with meditation the minimum observational change necessary to instill
an interest, to keep practicing.
